---
title: "Human Activity Recognition Using Smartphones Dataset Analysis Codebook"
author: "Elbern D. Sioquim"
date: "31 May 2020"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Code book for Analyzing the Human Activity Recognition Using Smartphones Data Set

This is a markdown document that describes how the Human Activity Recognition Using Smartphones Data Set was processed in order
to create a new tidy data set that calculates the average of the mean and standard deviation features that were extracted from
the original data set. The calculated average was based on the subject and activity performed by each subject which is shown in
the summary of the final data set.

```{r echo=FALSE, message=FALSE, warning=FALSE}
create_data_set <- function(type) {
    
    startFilePath <- "getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/"
    fileExtension <- ".txt"
    
    ## read activity data set
    filePath <- paste0(startFilePath, type, "/y_", type, fileExtension)
    activityDataSet <- read.table(filePath, header=FALSE, col.names = "activityCode")
    
    ## left outer join activity labels with activityDataSet to match the activity labels
    library(plyr)
    activityDataSetWithLabel <- join(activityDataSet,activityLabelsTable)
    
    ## read subject data set
    filePath <- paste0(startFilePath, type, "/subject_", type, fileExtension)
    subjectDataSet <- read.table(filePath, header=FALSE, col.names = "subject")
    
    ## read main data set and add column names in the data set based on features
    filePath <- paste0(startFilePath, type, "/X_", type, fileExtension)
    humanActivityRecognitionDataSet <- read.table(filePath,
                                                  header=FALSE,
                                                  col.names = featuresTable$featureName,
                                                  check.names=FALSE)
    
    ## bind acitivity name into the main data set
    humanActivityRecognitionDataSet <- cbind(activityName = activityDataSetWithLabel$activityName , humanActivityRecognitionDataSet)
    
    ## bind subject code into the main data set
    humanActivityRecognitionDataSet <- cbind(subject = subjectDataSet$subject , humanActivityRecognitionDataSet)
    
}
```

## 1. Read the activity labels data and add variable names for the columns
```{r echo=FALSE, message=FALSE, warning=FALSE}
## read acticvity labels
activityLabelsTable <- read.table("getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/activity_labels.txt",
                                  header=FALSE,
                                  sep = "",
                                  col.names = c("activityCode","activityName"))
print(activityLabelsTable)
```

## 2. Replace the underscore characters with space and convert the activity name to lowercase
```{r echo=FALSE, message=FALSE, warning=FALSE}
## convert to lowercase and replace special characters
activityLabelsTable$activityName <- tolower(activityLabelsTable$activityName)
activityLabelsTable$activityName <- gsub('_', ' ', activityLabelsTable$activityName)

print(activityLabelsTable)
```

## 3. Read the features info data into a data frame
```{r echo=FALSE, message=FALSE, warning=FALSE}
## read features
featuresTable <- read.table("getdata_projectfiles_UCI HAR Dataset/UCI HAR Dataset/features.txt",
                            header=FALSE,
                            sep = "",
                            col.names = c("featureIndex","featureName"))

print(featuresTable)
```

## 4. Read and create the data sets for test and train and include the subject and activityName variables.

A join was used to set the activityName variable using the activity labels and was column-binded to the
test or train data set. For the subject variable, it was directly column-binded to the data set since we
only need the subject code.

```{r echo=FALSE, message=FALSE, warning=FALSE}
## read and create the test and train data sets
humanActivityRecognitionTestDataSet <- create_data_set("test")
```
## Test data set
```{r echo=FALSE, message=FALSE, warning=FALSE}
str(humanActivityRecognitionTestDataSet)
humanActivityRecognitionTrainDataSet <- create_data_set("train")
```
## Train data set
```{r echo=FALSE, message=FALSE, warning=FALSE}
str(humanActivityRecognitionTrainDataSet)
```

## 5. Merge the test and train data sets and remove special characters from the variable names
```{r echo=FALSE, message=FALSE, warning=FALSE}
## merge the test and train data sets
mergedTestAndTrainDataSets <- rbind(humanActivityRecognitionTestDataSet,
                                    humanActivityRecognitionTrainDataSet)

## tidy column names
names(mergedTestAndTrainDataSets) <- tolower(names(mergedTestAndTrainDataSets))
names(mergedTestAndTrainDataSets) <- gsub("[^[:alnum:]]","",names(mergedTestAndTrainDataSets))

str(mergedTestAndTrainDataSets)
```

## 6. Create a new data set which would include the mean and standard deviation of the features
```{r echo=FALSE, message=FALSE, warning=FALSE}
## extract the mean and standard deviation for each measurement
mergedMeanAndStdDevDataSet <- mergedTestAndTrainDataSets[,grepl("subject|activity|mean|std", names(mergedTestAndTrainDataSets))]

str(mergedMeanAndStdDevDataSet)
```

## 7. Create a new data set by calculating the mean of the data based on activity and subject
```{r echo=FALSE, message=FALSE, warning=FALSE}
## create a new data by calculating the mean of the data based on activity and subject
library(dplyr)
meanBasedOnActivtyAndSubjectDataSet <- mergedMeanAndStdDevDataSet %>% group_by(subject, activityname) %>% summarise_all(mean)
```

## Output the new tidy data set into a text file
```{r echo=FALSE, message=FALSE, warning=FALSE}
## output to file the final data set containing the calculated data
write.table(meanBasedOnActivtyAndSubjectDataSet, file = "meanBasedOnActivtyAndSubjectDataSet.txt", sep = "\t", row.names = FALSE, col.names = TRUE)
print("meanBasedOnActivtyAndSubjectDataSet.txt")
```
## Dimensions of final data set
```{r echo=FALSE, message=FALSE, warning=FALSE}
textOutput <- paste("Number of observations :", nrow(meanBasedOnActivtyAndSubjectDataSet))
print(textOutput)
textOutput <- paste("Number of variables :", ncol(meanBasedOnActivtyAndSubjectDataSet))
print(textOutput)
```

## Structure of final data set
```{r echo=FALSE, message=FALSE, warning=FALSE}
str(meanBasedOnActivtyAndSubjectDataSet)
```

## Summary of final data set

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(meanBasedOnActivtyAndSubjectDataSet)
```

